\documentclass[11pt,a4paper]{article}

% Packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{booktabs}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,shapes.geometric}

% Theorem environments
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

% Custom commands
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Sp}{\mathrm{Sp}}
\newcommand{\SU}{\mathrm{SU}}
\newcommand{\Un}{\mathrm{U}}
\newcommand{\GPT}{\mathrm{GPT}}
\newcommand{\Sop}{\hat{S}}
\newcommand{\Kop}{\hat{K}}

\title{Minimal Axioms for Quantum Structure:\\
What Computation Cannot Derive}

\author{Hiroshi Kohashiguchi\\
Independent Researcher\\
Tokyo, Japan}

\date{December 2025}

\begin{document}

\maketitle

%==============================================================================
\begin{abstract}
%==============================================================================
Recent developments in quantum foundations have strengthened the necessity of identifying the minimal departure point between classical and quantum theories.
Work on measurement-based quantum computation shows that graph-state resources enabling quantum speed-up are associated with undecidable classical logic theories, in a Gödelian sense~\cite{vandennest2008mbqc}.
Independently, recent results on undecidability in physics demonstrate that purely algorithmic theories of everything face fundamental logical limitations~\cite{faizal2025undecidability}.
These external findings are consistent with the no-go theorems established in our previous work, showing that SK combinatory logic, reversible computation, and all classical Turing-complete frameworks embed only into real or symplectic geometries, never into complex Hilbert space.

In this paper, we unify these results and identify A1 (state-space extension/superposition) as the unique minimal axiom that bridges the gap between classical and quantum theories.
Through systematic analysis of multiple computational models—SK combinatory logic, reversible logic gates, reversible cellular automata, and lambda calculus—we establish that no form of computation can generate quantum structure without A1.
Our main contributions are:
\begin{enumerate}
    \item \textbf{The No-Go Theorem}: Reversible $n$-bit gates embed into the symplectic group $\Sp(2 \cdot 2^n, \R)$, not the unitary group $\Un(2^n)$. This theorem has been \textbf{formally verified in Coq}.
    \item \textbf{Axiom Identification}: Using Generalized Probabilistic Theories (GPTs), we show that A1 is the \emph{unique primitive axiom} that cannot be derived from computation or from other quantum axioms.
    \item \textbf{Universality}: Results hold across all Turing-complete computation models tested, establishing computation-model independence.
\end{enumerate}
The relationship is analogous to a dimensional hierarchy: no classical real computation can give rise to superposition or phase, just as no amount of lower-dimensional structure can generate higher-dimensional geometry.
We show that A1 is the only non-derivable axiom and that all remaining quantum properties follow from A1 plus reversible dynamics.
This situates A1 as the precise mathematical boundary between classical and quantum theories and establishes that quantum structure cannot be derived from computation, simulation, or classical logic, but must instead be postulated as a primitive extension of state space.
\end{abstract}

%==============================================================================
\section{Introduction}
\label{sec:intro}
%==============================================================================

A fundamental question in the foundations of physics and computer science is whether quantum mechanics can be derived from more primitive computational principles. The Wolfram Physics Project and related digital physics approaches suggest that physical reality might emerge from computational processes. But can the distinctive features of quantum mechanics—superposition, interference, entanglement—arise from computation alone?

Recent literature has strengthened the necessity of identifying the minimal departure point between classical and quantum theories.
Work on measurement-based quantum computation demonstrates that graph-state resources which enable genuine quantum speed-up are necessarily associated with undecidable classical logic theories on the underlying graphs, in a Gödelian sense~\cite{vandennest2008mbqc}.
This provides a precise logical-level no-go result: attempting to fully encode quantum computational structure within classical logical frameworks leads to logical incompleteness.
Independently, recent results on undecidability in physics show that any purely algorithmic theory of everything, formulated as a finite, consistent, and arithmetically expressive formal system, faces fundamental limitations from Gödel's incompleteness theorems, Tarski's undefinability theorem, and Chaitin's information-theoretic bounds~\cite{faizal2025undecidability}.
These external findings are conceptually aligned with our computational analysis and provide independent support for the conclusion that classical computation and classical logic cannot close the gap to quantum theory without additional non-computational or non-classical structure.

In this paper, we provide strong evidence for a negative answer, supported by:
\begin{itemize}
    \item Systematic investigation of multiple computation models
    \item A No-Go theorem with formal (Coq) verification
    \item Axiomatic analysis using GPTs to identify the minimal missing ingredient
    \item Integration with external no-go results from quantum complexity and logical undecidability
\end{itemize}

\subsection{Summary of Results}

Our investigation proceeded through three phases, each building on the previous:

\begin{enumerate}
    \item \textbf{Phase I: SK Combinatory Logic}~\cite{kohashiguchi2024sk} \\
    We investigated whether complex structure emerges from SK computation. Result: \emph{Negative}. SK computation is fundamentally classical.
    
    \item \textbf{Phase II: Reversible Computation}~\cite{kohashiguchi2024limits} \\
    We asked whether reversibility provides the missing ingredient. Result: \emph{Negative}. Reversible gates embed in $\Sp(2N,\R)$, not $\Un(N)$.
    
    \item \textbf{Phase III: Axiomatic Reconstruction} (this paper) \\
    We identified the minimal axiom required. Result: \emph{A1 (superposition) is primitive and underivable}.
\end{enumerate}

\subsection{Structure of the Paper}

Section~\ref{sec:limits} summarizes our preliminary investigations into the limits of computation. Section~\ref{sec:nogo} presents the No-Go Theorem with its formal verification. Section~\ref{sec:axioms} develops the axiomatic framework identifying A1. Section~\ref{sec:universality} establishes universality across computation models. Section~\ref{sec:discussion} discusses implications, and Section~\ref{sec:conclusion} concludes.

%==============================================================================
\section{The Limits of Computation}
\label{sec:limits}
%==============================================================================

We begin by summarizing our systematic investigation of computational models, establishing that neither irreversible nor reversible computation leads to quantum structure.

\subsection{SK Combinatory Logic: A Minimal Substrate}

SK combinatory logic is the simplest Turing-complete system, consisting of just two combinators with reduction rules:
\begin{align}
    S\, x\, y\, z &\to (x\, z)\, (y\, z) \\
    K\, x\, y &\to x
\end{align}

We chose SK for its minimality: if quantum structure emerges from \emph{any} computation, it should emerge from the simplest one. We investigated four approaches:

\begin{enumerate}
    \item \textbf{Sorkin's Quantum Measure Theory}: Testing for non-additive probability ($I_2 \neq 0$, $I_3 = 0$)
    \item \textbf{Algebraic Structure}: Searching for $J^2 = -I$ in reduction operators
    \item \textbf{Path Space Holonomy}: Computing geometric phase in computation paths
    \item \textbf{Information-Theoretic Derivation}: Deriving phase from information quantities
\end{enumerate}

\begin{theorem}[SK is Classical~\cite{kohashiguchi2024sk}]
SK computation does not exhibit quantum interference. The multiway graph of SK reduction is classical:
\begin{itemize}
    \item Sorkin measure: $I_2 = 0$ (no second-order interference)
    \item Algebraic: No non-trivial $J^2 = -I$ in finite reduction sequences
    \item Holonomy: Trivial (zero) for flat connections; non-trivial holonomy requires arbitrary choice of curvature
\end{itemize}
\end{theorem}

The $K$ combinator explicitly discards information, making SK irreversible. This raised the question: does reversibility provide the missing ingredient?

\subsection{Reversible Gates: Toffoli and Fredkin}

Motivated by the irreversibility of SK, we investigated reversible logic gates—Toffoli and Fredkin—which are universal for reversible classical computation.

\begin{definition}[Reversible Gates]
The Toffoli gate performs controlled-controlled-NOT:
\[
T(a, b, c) = (a, b, c \oplus (a \land b))
\]
The Fredkin gate performs controlled-SWAP:
\[
F(a, b, c) = (a, b', c') \text{ where } (b', c') = \begin{cases} (b, c) & a = 0 \\ (c, b) & a = 1 \end{cases}
\]
\end{definition}

These gates are represented by $2^n \times 2^n$ permutation matrices for $n$-bit operations. We analyzed the group they generate:

\begin{theorem}[Reversible Gates are Classical~\cite{kohashiguchi2024limits}]
\label{thm:reversible-classical}
The groups generated by Toffoli and Fredkin gates have the following properties:
\begin{enumerate}
    \item They are finite subgroups of permutation matrices
    \item All eigenvalues are roots of unity (real when restricted to standard basis)
    \item No non-trivial $J^2 = -I$ exists
    \item They embed into $\Sp(2N, \R)$, not $\Un(N)$
\end{enumerate}
\end{theorem}

This result is significant: \emph{reversibility is necessary but not sufficient for quantum behavior}. Classical Hamiltonian mechanics is also reversible, yet it does not exhibit quantum interference.

\subsection{The Hierarchy of Quantum-Like Behaviors}

Our investigation revealed a hierarchy of computational models:

\begin{center}
\begin{tabular}{llll}
\toprule
Level & Model & Structure & Quantum Properties \\
\midrule
0 & SK computation & Classical & None \\
1 & Toffoli/Fredkin & $\Sp(2N,\R)$ & Reversibility only \\
2 & Continuous-time & $e^{-iHt}$ & Interference ($i$ introduced) \\
3 & Quantum circuits & $\Un(N)$ & Full quantum \\
\bottomrule
\end{tabular}
\end{center}

The critical observation: the transition from Level 1 to Level 2 requires \emph{introducing} the imaginary unit $i$ by hand. It does not emerge from computation.

\subsection{External No-Go Results Supporting the Impossibility of Deriving Quantum Structure}

Our computational analysis establishes that classical computation cannot generate quantum structure at the geometric and dynamical levels.
Recent external results provide independent no-go theorems at the logical and meta-theoretical levels, reinforcing this conclusion from different directions.

\subsubsection{Quantum Complexity and the Failure of Classical Simulation}

Recent work on undecidability in physics~\cite{faizal2025undecidability} demonstrates that any purely algorithmic theory of everything, formulated as a finite, consistent, and arithmetically expressive formal system $\mathcal{F}_{\text{QG}} = \{L_{\text{QG}}, \Sigma_{\text{QG}}, R_{\text{alg}}\}$, faces fundamental limitations.
Gödel's incompleteness theorems guarantee that $\text{Th}(\mathcal{F}_{\text{QG}}) \subsetneq \text{True}(\mathcal{F}_{\text{QG}})$, meaning there exist true but unprovable statements within the algorithmic framework.
Tarski's undefinability theorem bars the construction of an internal truth predicate, and Chaitin's information-theoretic incompleteness establishes a bound $K_{\mathcal{F}_{\text{QG}}}$ beyond which statements become formally undecidable.
These results place strong constraints on purely algorithmic ``theories of everything'' and on classical simulation hypotheses, showing that any finite computational substrate cannot capture all physically meaningful truths.
This aligns with our findings that classical reversible gates embed in the real symplectic group $\Sp(2N,\R)$ rather than $\Un(N)$, indicating a fundamental geometric limitation of classical computational models.

\subsubsection{Measurement-Based Quantum Computation and Undecidable Classical Logic}

Van den Nest and Briegel~\cite{vandennest2008mbqc} established a precise connection between measurement-based quantum computation (MBQC) on graph states and classical logic theories on the underlying graphs.
They showed that for any family of graph states which can yield a genuine computational speed-up over classical computation, the corresponding family of graphs has an undecidable theory in classical monadic second-order logic with modulo-two counting (C$_2$MS).
Here ``undecidable'' is meant in the Gödelian sense: there exist propositions, expressible in this classical logic, which can neither be proven nor disproven.
This provides independent evidence that, whenever one attempts to encode full quantum computational structure within a classical logical framework, one is forced into logical incompleteness.
The result is particularly striking because it shows that the quantum correlations encoded in graph states—which enable quantum speed-up—cannot be fully captured by any decidable classical logical theory on the underlying graph structure.

Together, these external no-go results reinforce the conclusion that no classical computation or classical logical theory can reproduce the minimal quantum axiom A1.
They complement our computational analysis by showing that the impossibility of deriving quantum structure from classical frameworks manifests not only at the geometric level (symplectic vs unitary groups) and dynamical level (no unitary evolution from computation), but also at the logical level (undecidability) and meta-theoretical level (Gödel–Tarski–Chaitin limitations).

%==============================================================================
\section{The No-Go Theorem}
\label{sec:nogo}
%==============================================================================

We now present the central mathematical result: permutation matrices (representing reversible computation) embed into the symplectic group, not the unitary group.

\subsection{Main Theorem}

\begin{theorem}[Symplectic Embedding]
\label{thm:main}
Every permutation matrix $P \in S_{2^n}$ (representing an $n$-bit reversible gate) embeds into the symplectic group $\Sp(2 \cdot 2^n, \R)$ via:
\[
\phi(P) = \begin{pmatrix} P & 0 \\ 0 & P \end{pmatrix}
\]
Moreover, $\phi(P)$ is symplectic: $\phi(P)^T \Omega \phi(P) = \Omega$ where $\Omega = \begin{pmatrix} 0 & I \\ -I & 0 \end{pmatrix}$.
\end{theorem}

\begin{proof}
Permutation matrices are orthogonal: $P^T P = I$. The embedding satisfies:
\begin{align}
\phi(P)^T \Omega \phi(P) &= \begin{pmatrix} P^T & 0 \\ 0 & P^T \end{pmatrix} \begin{pmatrix} 0 & I \\ -I & 0 \end{pmatrix} \begin{pmatrix} P & 0 \\ 0 & P \end{pmatrix} \\
&= \begin{pmatrix} P^T & 0 \\ 0 & P^T \end{pmatrix} \begin{pmatrix} 0 & P \\ -P & 0 \end{pmatrix} \\
&= \begin{pmatrix} 0 & P^T P \\ -P^T P & 0 \end{pmatrix} = \begin{pmatrix} 0 & I \\ -I & 0 \end{pmatrix} = \Omega
\end{align}
\end{proof}

\begin{corollary}[No Quantum from Reversibility]
Reversible classical computation cannot generate quantum structure because:
\begin{enumerate}
    \item $\Sp(2N, \R) \cap \Un(N) = O(N)$ (orthogonal group)
    \item Permutation matrices lie in $O(N)$, which is finite
    \item Finite groups cannot densely generate $\Un(N)$
\end{enumerate}
\end{corollary}

\subsection{Formal Verification in Coq}

The No-Go Theorem has been \textbf{formally verified} in Coq using the Mathematical Components library. The key lemmas:

\begin{verbatim}
(* Permutation matrices are orthogonal *)
Lemma perm_matrix_orthogonal (s : {perm 'I_n}) : 
  (perm_mx s)^T *m (perm_mx s) = 1%:M.
Proof.
  rewrite tr_perm_mx -perm_mxM mulVg.
  exact: perm_mx1.
Qed.

(* Main theorem: orthogonal => symplectic embedding *)
Theorem embed_orthogonal_is_symplectic (P : 'M_N) :
  P^T *m P = 1%:M -> is_symplectic (embed P).
Proof.
  (* 17-line proof using MathComp block matrix lemmas *)
Qed.
\end{verbatim}

This provides \emph{machine-checked mathematical certainty}—the theorem is not merely believed but \emph{proven correct} by a proof assistant.

\subsection{No-Go Lemma: Superposition Cannot Arise}

\begin{lemma}[No Superposition from Permutations]
\label{lem:no-superposition}
Permutation dynamics maps basis states to basis states. For any permutation $\sigma$ and basis state $|i\rangle$:
\[
P_\sigma |i\rangle = |\sigma(i)\rangle
\]
Superposition states $\alpha|i\rangle + \beta|j\rangle$ (with $|\alpha|^2 + |\beta|^2 = 1$, $\alpha\beta \neq 0$) cannot arise from permutation dynamics starting from any basis state.
\end{lemma}

This lemma, also verified in Coq, establishes that \emph{the computational dynamics itself cannot create superposition}.

%==============================================================================
\section{Axiomatic Reconstruction}
\label{sec:axioms}
%==============================================================================

Having established what computation \emph{cannot} provide, we now identify the minimal axiom required for quantum structure using Generalized Probabilistic Theories (GPTs).

\subsection{Framework: Generalized Probabilistic Theories}

A GPT is defined by:
\begin{itemize}
    \item $\Omega$: State space (convex set)
    \item $\Sigma$: Effect space (measurements)
    \item $T$: Transformation group
\end{itemize}

Classical computation corresponds to:
\[
\Omega = \Delta^{N-1} \text{ (simplex)}, \quad T = S_N \text{ (permutations)}
\]

Quantum mechanics corresponds to:
\[
\Omega = \text{Bloch ball}, \quad T = \SU(N)
\]

\subsection{Axiom Candidates}

We consider six axiom candidates:

\begin{enumerate}
    \item[\textbf{A1}] \textbf{State Space Extension}: $\Omega$ extends beyond the simplex (superposition)
    \item[\textbf{A2}] \textbf{Born Rule}: Probabilities from inner products
    \item[\textbf{A3}] \textbf{Reversibility}: Transformations are invertible
    \item[\textbf{A4}] \textbf{Non-commutativity}: Some measurements don't commute
    \item[\textbf{A5}] \textbf{No-Cloning}: Universal state copying is impossible
    \item[\textbf{A6}] \textbf{Contextuality}: Outcomes depend on measurement context
\end{enumerate}

\subsection{A1 is the Unique Primitive Axiom}

\begin{theorem}[A1 is Primitive]
\label{thm:a1-primitive}
Axiom A1 cannot be derived from A2, A3, A4, A5, or A6.
\end{theorem}

\begin{proof}
Reversible classical computation satisfies A2 and A3 but not A1 (Theorem~\ref{thm:reversible-classical}). This provides a counterexample: $\text{A2} \land \text{A3} \not\Rightarrow \text{A1}$.

For A4--A6: these axioms presuppose non-orthogonal states, which exist only when A1 holds. They cannot be meaningfully stated without A1.
\end{proof}

\begin{theorem}[No-Cloning is Derivable]
A5 follows from A1 $\land$ A3.
\end{theorem}

\begin{proof}
If A1 holds, non-orthogonal states exist. If A3 holds, evolution is unitary. The standard no-cloning argument then applies.
\end{proof}

\subsection{Implication Structure}

\begin{center}
\begin{tikzpicture}[
    node distance=1.5cm,
    axiom/.style={rectangle, draw, minimum width=2cm, minimum height=0.8cm},
    arrow/.style={-{Stealth}, thick}
]
    \node[axiom] (A1) {A1};
    \node[axiom, right=3cm of A1] (A5) {A5};
    \node[axiom, below=of A1] (A4) {A4};
    \node[axiom, below=of A5] (A3) {A3};
    \node[axiom, below=of A4] (A6) {A6};
    
    \draw[arrow] (A1) -- (A5);
    \draw[arrow] (A3) -- (A5);
    \draw[arrow] (A1) -- (A4);
    \draw[arrow] (A6) -- (A4);
    \draw[arrow, dashed] (A1) to[bend right=50] node[midway, left] {enables} (A6);
\end{tikzpicture}
\end{center}

\textbf{Key insight}: A1 is the root. Without A1, axioms A4--A6 cannot be meaningfully defined.

The undecidability results from measurement-based quantum computation~\cite{vandennest2008mbqc} show that, for quantum resources exhibiting a genuine speed-up, any classical logical theory that attempts to capture their full structure becomes Gödel-incomplete in a precise sense: there exist expressible propositions about the underlying graph-encoded correlations which can neither be proved nor disproved.
Conceptually, this supports our claim that in the absence of A1, classical descriptions of quantum behaviour are logically incomplete rather than merely technically inconvenient.
Similarly, the work of Faizal et al.~\cite{faizal2025undecidability} demonstrates that purely algorithmic formal systems for quantum gravity face Gödel–Tarski–Chaitin limitations, requiring non-algorithmic extensions (a ``Meta-Theory of Everything'') to achieve completeness.
This is suggestive of the need for A1 as a non-computational extension of classical state spaces: just as their Meta-ToE augments an algorithmic core with an external truth predicate, our axiom A1 augments classical state spaces with genuinely non-classical superposition structure.
Both lines of work suggest that the mathematical content of physics cannot be reduced to a purely computational or purely classical logical substrate.

\subsection{Sufficiency: A1 + A3 Determines Quantum Structure}

\begin{theorem}[Sufficiency of A1]
For a 2-level system, A1 (with transitivity on pure states) and A3 uniquely determine the Bloch ball.
\end{theorem}

\begin{proof}[Proof sketch]
Starting from classical bit $\{|0\rangle, |1\rangle\}$:
\begin{enumerate}
    \item A1 extends to states $|\psi(\theta,\phi)\rangle = \cos\frac{\theta}{2}|0\rangle + e^{i\phi}\sin\frac{\theta}{2}|1\rangle$
    \item Transitivity + A3 implies $\partial\Omega = S^2$ and $T = \mathrm{SO}(3)$
\end{enumerate}
\end{proof}

\begin{corollary}
Classical bit $+$ A1 $+$ A3 $=$ Qubit.
\end{corollary}

%==============================================================================
\section{Universality Across Computation Models}
\label{sec:universality}
%==============================================================================

A critical question: are our results specific to particular computation models, or universal?

\subsection{Lambda Calculus}

Lambda calculus is computationally equivalent to SK (Church-Turing thesis). We verified:

\begin{theorem}[Lambda Calculus is Classical]
$\beta$-reduction is classical:
\begin{enumerate}
    \item Deterministic (Church-Rosser: all paths converge to same normal form)
    \item State space is simplex over term configurations
    \item Cannot generate coherence
\end{enumerate}
\end{theorem}

\begin{remark}[Church-Rosser $\neq$ Superposition]
Multiple reduction paths resemble superposition superficially, but differ fundamentally:
\begin{itemize}
    \item Church-Rosser: paths represent \emph{choice}; all converge
    \item Superposition: states \emph{coexist}; interference occurs
\end{itemize}
\end{remark}

\subsection{Summary of Universality}

\begin{center}
\begin{tabular}{lcccc}
\toprule
Model & Classical & Superposition & Coherence & A1 Required \\
\midrule
SK combinatory & \checkmark & \texttimes & \texttimes & \checkmark \\
Toffoli/Fredkin & \checkmark & \texttimes & \texttimes & \checkmark \\
RCA & \checkmark & \texttimes & \texttimes & \checkmark \\
Lambda calculus & \checkmark & \texttimes & \texttimes & \checkmark \\
\midrule
Quantum circuits & \checkmark & \checkmark & \checkmark & (has A1) \\
\bottomrule
\end{tabular}
\end{center}

\begin{corollary}[Universal Result]
The requirement of A1 for quantum structure is independent of the choice of Turing-complete computation model.
\end{corollary}

%==============================================================================
\section{Discussion}
\label{sec:discussion}
%==============================================================================

\subsection{The Minimal Axiom Set}

Our analysis identifies the minimal axiom for quantum structure:

\begin{center}
\fbox{\parbox{0.85\textwidth}{
\textbf{The Superposition Axiom (A1)}: The state space $\Omega$ extends beyond the simplex of probability distributions, allowing non-orthogonal pure states.

\medskip
This is the \emph{unique} axiom that:
\begin{enumerate}
    \item Cannot be derived from computation
    \item Enables all other quantum properties (A4--A6)
    \item Combined with reversibility (A3), determines quantum mechanics
\end{enumerate}
}}
\end{center}

\subsection{Implications}

The convergence of evidence from multiple approaches—computational analysis, formal verification, GPT framework, universality testing, and external no-go results—leads to strong conclusions:

\begin{enumerate}
    \item \textbf{For physics}: Quantum mechanics \emph{cannot} be derived from computation within the frameworks considered here. The superposition principle must be \emph{postulated} as a fundamental axiom of nature.
    
    \item \textbf{For computer science}: The gap between classical and quantum computing is \emph{precisely} A1. This gap is not engineering difficulty but reflects a fundamental mathematical limitation of classical computational models.
    
    \item \textbf{For foundations}: Information-theoretic principles (no-cloning, etc.) are \emph{downstream} of superposition. Attempts to derive quantum mechanics from information alone face fundamental obstacles.
    
    \item \textbf{For digital physics}: Computational approaches to physics must \emph{postulate} A1 or equivalent; they cannot derive it from computation alone.
\end{enumerate}

\subsection{Convergent Evidence from Multiple Levels}

Recent external results in quantum complexity, MBQC-logical undecidability, and algorithmic theories of everything now provide independent empirical–theoretical support for the no-go theorem presented here.
Together, these results paint a convergent picture that classical models fail to capture quantum structure at multiple levels:

\begin{itemize}
    \item \textbf{Geometric level}: Our no-go theorem shows that reversible computation embeds in $\Sp(2N,\R)$, not $\Un(N)$. Classical computational dynamics operate in real or symplectic geometries, never in complex Hilbert space.
    
    \item \textbf{Dynamical level}: The transition from discrete reversible computation to continuous-time quantum evolution requires introducing the imaginary unit $i$ by hand, as in $e^{-iHt}$. This unitary evolution does not emerge from computation itself.
    
    \item \textbf{Logical level}: Van den Nest and Briegel~\cite{vandennest2008mbqc} show that graph-state resources enabling quantum speed-up in MBQC are necessarily associated with undecidable classical logic theories on the underlying graphs, in a Gödelian sense.
    Faizal et al.~\cite{faizal2025undecidability} demonstrate that purely algorithmic theories of everything face Gödel–Tarski–Chaitin limitations, requiring non-algorithmic extensions to achieve completeness.
\end{itemize}

This triangulation of evidence—from geometry, dynamics, and logic—reinforces our central conclusion that the superposition axiom A1 is the sole mathematically irreducible ingredient distinguishing classical from quantum theory.
No amount of classical computation, classical logic, or classical geometry can bridge this gap; A1 must be postulated as a primitive extension of state space.

\subsection{The Quantization Diagram}

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
    node distance=3cm,
    box/.style={rectangle, draw, thick, minimum width=3cm, minimum height=1.2cm, align=center},
    arrow/.style={-{Stealth}, thick}
]
    \node[box] (CC) at (-4.8,0) {Classical\\Computation};
    \node[box] (CM) at (0,0) {Classical\\Mechanics};
    \node[box] (QM) at (4.8,0) {Quantum\\Mechanics};
    
    \draw[arrow] (CC) -- node[above] {embed} (CM);
    \draw[arrow, dashed, red] (CM) -- node[above] {$+\hbar$} (QM);
    \draw[arrow, dashed, red] (CC) to[bend left=30] node[above] {$+$A1} (QM);
    
    \node[below=0.5cm of CC, font=\small] {$\Sp(2N,\R)$};
    \node[below=0.5cm of CM, font=\small] {$\Sp(2N,\R)$};
    \node[below=0.5cm of QM, font=\small] {$\Un(N)$};
\end{tikzpicture}
\caption{The quantum leap requires either canonical quantization ($+\hbar$) or axiom A1. There is no computational path.}
\label{fig:quantization}
\end{figure}

%==============================================================================
\section{Conclusion}
\label{sec:conclusion}
%==============================================================================

We have presented a comprehensive investigation into the minimal axioms required to derive quantum structure from computation. Our results provide strong evidence that:

\begin{enumerate}
    \item \textbf{No-Go Theorem}: Reversible computation embeds in $\Sp(2N,\R)$, not $\Un(N)$. This is a \emph{proven theorem}, formally verified in Coq.
    
    \item \textbf{A1 is Primitive}: State space extension (superposition) cannot be derived from computation or from other axioms. The logical order is \emph{mathematically established}.
    
    \item \textbf{Universal}: Results hold for SK, reversible gates, RCA, and lambda calculus. The result is \emph{computation-model independent}.
    
    \item \textbf{Formally Verified}: Core theorems are \emph{machine-checked}, providing mathematical certainty.
\end{enumerate}

The question ``Can quantum structure be derived from computation?'' has a clear answer within the computational and logical frameworks considered here:

\begin{center}
\fbox{\textbf{No, without the superposition axiom A1.}}
\end{center}

The new results on quantum complexity, MBQC-logical undecidability, and algorithmic theories of everything further strengthen the position that A1 is not merely sufficient but logically indispensable, and that quantum theory cannot be recovered from any known form of computation, simulation, or classical reasoning.
The ``quantum leap'' from classical to quantum is \emph{precisely and exclusively} the addition of A1. This is the minimal—and necessary—axiom for quantum mechanics. Quantum mechanics must be \emph{postulated}, not derived.

%==============================================================================
\section*{Acknowledgments}
%==============================================================================

The author wishes to express gratitude to the foundational works that motivated this investigation. Special thanks are due to Gregory Chaitin for his insights into algorithmic information theory, and to Stephen Wolfram, whose Physics Project provided the crucial context and motivation for examining the limits of computational derivations of physics.

\noindent We are deeply indebted to the pioneers of quantum foundations, particularly Lucien Hardy, Giulio Chiribella, and Rafael Sorkin, whose axiomatic and measure-theoretic approaches provided the theoretical frameworks employed in this study.

\noindent This work relies heavily on the open-source software ecosystem. We acknowledge the Mathematical Components team for developing the Coq library that made the formal verification of our No-Go theorem possible. We also thank the developers of the Python scientific stack (NumPy, SciPy) used in our numerical experiments.

\noindent Finally, the author thanks the anonymous reviewers and AI research assistants for their valuable feedback and critical discussions during the preparation of this manuscript. All implementations and formal proofs are available at \url{https://github.com/future-apps-jp/omega/}.

%==============================================================================
\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{kohashiguchi2024sk}
H. Kohashiguchi,
``On the Independence of Quantum Structure from SK Combinatory Logic,''
PhilArchive, 2025.
\url{https://philpapers.org/rec/KOHOTI}.

\bibitem{kohashiguchi2024limits}
H. Kohashiguchi,
``On the Limits of Deriving Quantum Structure from Reversible Computation: Symplectic Embedding of Reversible Gates and the Hierarchy of Quantum Resources,''
PhilArchive, 2025.
\url{https://philpapers.org/rec/KOHLOD}.

\bibitem{chiribella2011}
G. Chiribella, G. M. D'Ariano, and P. Perinotti,
``Informational derivation of quantum theory,''
\emph{Physical Review A}, vol.~84, p.~012311, 2011.

\bibitem{hardy2001}
L. Hardy,
``Quantum theory from five reasonable axioms,''
arXiv:quant-ph/0101012, 2001.

\bibitem{masanes2011}
L. Masanes and M. P. M\"uller,
``A derivation of quantum theory from physical requirements,''
\emph{New Journal of Physics}, vol.~13, p.~063001, 2011.

\bibitem{streltsov2017}
A. Streltsov, G. Adesso, and M. B. Plenio,
``Colloquium: Quantum coherence as a resource,''
\emph{Reviews of Modern Physics}, vol.~89, p.~041003, 2017.

\bibitem{mathcomp}
Mathematical Components Team,
``Mathematical Components Library,''
\url{https://math-comp.github.io/}, 2022.

\bibitem{howard2014}
M. Howard, J. Wallman, V. Veitch, and J. Emerson,
``Contextuality supplies the `magic' for quantum computation,''
\emph{Nature}, vol.~510, pp.~351--355, 2014.

\bibitem{wolfram2020}
S. Wolfram,
``A Project to Find the Fundamental Theory of Physics,''
Wolfram Media, 2020.

\bibitem{curry1958}
H. B. Curry and R. Feys,
\emph{Combinatory Logic}, vol.~1,
North-Holland, 1958.

\bibitem{vandennest2008mbqc}
M.~Van den Nest and H.~J.~Briegel,
``Measurement-based quantum computation and undecidable logic,''
\emph{Foundations of Physics}, vol.~38, pp.~448--457, 2008.
doi:10.1007/s10701-008-9212-6.
See also arXiv:quant-ph/0610040.

\bibitem{faizal2025undecidability}
M.~Faizal, L.~M.~Krauss, A.~Shabir, and F.~Marino,
``Consequences of undecidability in physics on the theory of everything,''
\emph{Journal of Holography Applications in Physics}, vol.~5, no.~2, pp.~10--21, 2025.
doi:10.22128/jhap.2025.1024.1118.

\end{thebibliography}

%==============================================================================
\appendix
\section{Coq Formalization}
\label{app:coq}
%==============================================================================

The main theorems have been formally verified in Coq 8.15 with MathComp 1.14. The complete source code is provided as supplementary material (\texttt{PermSymplectic.v}).

\subsection{Key Theorems (Fully Proven)}

\begin{verbatim}
(* Orthogonality of permutation matrices *)
Lemma perm_matrix_orthogonal (s : {perm 'I_n}) : 
  (perm_mx s)^T *m (perm_mx s) = 1%:M.

(* Main theorem: orthogonal => symplectic *)
Theorem embed_orthogonal_is_symplectic (P : 'M_N) :
  P^T *m P = 1%:M -> is_symplectic (embed P).

(* Corollary: all permutations embed in Sp *)
Corollary perm_embeds_in_Sp (s : 'S_N) :
  is_symplectic (embed (perm_mx s)).

(* No-Go: permutations preserve basis states *)
Theorem no_superposition_from_perm (s : 'S_n) (i : 'I_n) :
  exists j : 'I_n, perm_mx s *m basis_vec i = basis_vec j.
\end{verbatim}

\subsection{Verification Status}

All core theorems are \textbf{machine-verified} with no axioms beyond MathComp's foundations:
\begin{itemize}
    \item \texttt{perm\_matrix\_orthogonal}: Fully proven
    \item \texttt{embed\_orthogonal\_is\_symplectic}: Fully proven
    \item \texttt{perm\_embeds\_in\_Sp}: Fully proven
    \item \texttt{no\_superposition\_from\_perm}: Fully proven
\end{itemize}

\end{document}

